---
title: "Datamanagement"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)
```


# Loading Dataset

```{r}
# Laden des Datensatzes
library(readstata13)
library(tidyverse)
library(magrittr)

#Daten als Objekt importieren
econ <- read.dta13(file="data/offline/econometrics.dta" ,
                   convert.factors=F,
                   nonint.factors = F)

# Subsample with 5000 Observations
econ <- econ [sample(c(1:nrow(econ)),size = 5000,replace = FALSE),]


```


# Recoding/Variable declaration

!! Marcel here: not sure if we should stick with car::recode. I personally don't like it and prefer classic R with [], but its so much more code-intensive...

```{r}

library(car)

# Variable Alter generieren
econ$alter = 2003 - econ$gebjahr
econ$alter [econ$alter == 2004] <- NA # entfernen der fehlerhaften

#sex 
econ$sex <- factor(econ$sex,levels = c(1,2),labels = c("männlich","weiblich"))

# Ueberstunden recodieren 0=nein, 1=ja
econ$over = recode(econ$tp72,"2=0;-2=NA;-1=NA; 3=NA")

# Vertragliche und tatsaechliche Wochenarbeitszeit
# Missings bereinigen
econ$contract = recode(econ$tp7001,"-3=NA;-2=NA;-1=NA")
econ$actual = recode(econ$tp7003,"-3=NA;-2=NA;-1=NA")
econ$contract = econ$contract/10
econ$actual = econ$actual/10

# Vertrauen 
# "Trust in people" und "Can't rely on anybody" recodieren
econ$trust = recode(econ$tp0301,"-1=NA")
econ$rely = recode(econ$tp0302,"-1=NA")

econ$netinc = recode(econ$tp7602,"-3=NA;-2=NA;-1=NA")

# Nur relevante Variablen werden übernommen
econ_data <- econ[,c("netinc","alter","sex","contract","actual","trust","rely")]
head(econ_data)

# Delete NAs
econ_data <- na.omit(econ_data)

# Einkommen Kategorisiert in Quartile

hist(econ$netinc)
quantile(econ_data$netinc)

econ_data$inc_kat <- NA
econ_data$inc_kat [econ_data$netinc < quantile(econ_data$netinc)[2]] <- "Q1"
econ_data$inc_kat [econ_data$netinc >= quantile(econ_data$netinc)[2] & econ_data$netinc < quantile(econ_data$netinc)[3]] <- "Q2"
econ_data$inc_kat [econ_data$netinc >= quantile(econ_data$netinc)[3] & econ_data$netinc < quantile(econ_data$netinc)[4]] <- "Q3"
econ_data$inc_kat [econ_data$netinc >= quantile(econ_data$netinc)[4] ] <- "Q4"

table(econ_data$inc_kat)
```


# Advanced DM: The tidyverse

Some of the most important functions when it gets to datamanagement stem from the tidyverse-packages and most notably dplyr. We can only provide you with a small overview here, but if you understand `mutate`, `summarise` and `group_by`, you should be good to go. 

## Mutate

We can easily add variables (or transformed versions of others) to a knew dataset by adding a new variable through mutate. 

For example, assume we wanted to add a squared age term: 

```{r}

econ_data <- 
  econ_data %>% mutate(alter_squared = alter^2)

head(econ_data)

```

## Summarise

Summarise allows us to summarise certain variables, such as certain features of netincome. 

```{r}

econ_data %>% 
  summarise(mean = mean(alter,na.rm = T),
            sd = sd(alter,na.rm = T))

```

## group_by

The same is also possibled for grouped structures. Say, for example, you would want to calculate seperate values for different genders: 

```{r}

econ_data %>% 
  group_by() %>% 
  summarise(mean = mean(alter,na.rm = T),
            sd = sd(alter,na.rm = T))

```
