---
title: "Objects and Data Frames"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




***
# File names
Be sure to save all your files (especially the data input) in the folders below where you have started your project.
To let R open files you have to use the forward slash (not the backward one).
Example:
```{r eval=FALSE}
a=read.csv2("data/SOEP.csv")
```

# Objects in R
## Basic Calculations
We will often watn to store reuslts of calculations to reuse them later. For this, we can work with basic objects. An object has a name and a conent. We can freely choose the name of an object givan certain rules - they have to start with a letter and include only letters, numbers and some süpecial characters (".", "_", "-"). **R is case sensitive so "x" and "X" are different object names**.
The content of an object is assigned using "<-" or "=".

In order to assign the value of 5 to the object econ type
```{r}
x = 5
# or
x <- 5
```
If you already before had an object named x and now create it again, the old version will be overwritten. Now you can use "x" in your calculations. 

Example:
```{r}
5*x
```

A list of currently ddefined object names can be obtained using ```ls() ```
In R Studio all the object names are also shown in the "Workspace" window on the top right side. 
```{r}
# Change and Delete objects:
rm(x) # Deletes an object
rm(list=ls()) # all objects are removed
```

***
## Vectors
For statistical calculations, we obviously need to work with data sets including many numbers of instead of scalears. The simplest way we can collect many numbers (or other tpyes of information) is called a vector in R terminology (you have already been familiarized with vectors and the page before). 

To define a vector, we can collect different values using ```c(value1, value2,...)```. 
```{r}
# Examples
a=c(1,2,3,4)
b=a+1
b
c=sqrt(b+a)
c
```

```{r}
# Important R functions for vectors:
# Define vector
(a <- c(7,2,6,9,4,1,3))

# Basic functions:
sort(a)
length(a)
min(a)
max(a)
sum(a)
prod(a)

# Creating special vectors:
numeric(20)
rep(1,20)
seq(50)
5:15
seq(4,20,2)
```

dsdfsdafsdfdsfs
# taschenrechneR
***

### ```x %in% y```	
Checks which elements of vector x are included in y
```{r}
1:3 %in% 2:4
```	


***
### ```# lorem ipsum```
Comment. Everything behind a hashtag is not run by the program 

```{r}
1+1 #Hier habe ich eins und eins zusammengezählt
```	
  
***
### ```y <- x```  
Attribute x to y
```{r}
a <- "hello world"
a
```	


***

### ```x:y```  
Generates sequence of ganzzahlen
```{r}
1:3
```	

***

***
# R algebra

***
## Classes 

***
### Vectors

numeric,character,factor

***
### Matrices

***
### Data Frames

***
### Lists


***
## Indexing

### ```x[.]```  
Indexing of a vector 
```{r}
a <- c(1,2,3)
names(a) <- c("q","r","s")
a[1]
a[c(TRUE,TRUE,FALSE)]
a["s"]
```	

### ```x[.,.]```  
Indexing of data.frames (matix)	
```{r}
data(iris)
iris[1,1]
iris[1:2,1:2]
iris[1,]
```	

***
### ```$```
data(iris)
```{r}
data(iris)
table(iris$Species)
```	
***
### ```x == y; ; x != y ,x < y; x > y; x <= y; x >= y  ```	
Logical operators. Return TRUE/FALSE VALUES.
```{r}
1 == 2 #Gleichheit
2 != 3 #Ungleichheit
2 < 3 #Kleiner
2 <= 3 #Kleiner gleich
```	
vectorisation possible
***
```{r}
1 == c(1,2,3)
```	


***

# Useful functions

### ```c()``` 
Combines elements to a vector
```{r}
c(1,2)
```



### ```read_spss(){haven}```
Read spss files via haven  (DIE HIER EHER DATAMANAGEMENT)
```
ess <- read_spss(file = "data/ESS7e02_1.sav",
                 user_na = TRUE)
```

***

### ```read_dta(){readstata13}``` 
Read dta files 
```
lala <- read_dta(file = "blabla",version = 13)
```

***

### ```within(data,{befehl})``` 
Execute r command within data environment. Returns modified dataset.```{}``` make sure that commands are executed across several lines.
```{r}
data(iris)
summary(iris$Sepal.Length)
iris <- within(iris,{
  Sepal.Length <- Sepal.Length + 10
})
summary(iris$Sepal.Length)
```

***

### ```table(x)``
***

### ```prop.table()```
***

### ```summary(x)```
***

### ```mean(x)```
***

### ```sum(x)```
***

### ```sd(x)```
***

### ```median(x)```

### ```round(2)```

### ```factor()```


